Stage-Splitting a Modal Language

### 2014-09-29 paper/semantics.tex
Small edits.

fun qs (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qs (left, k)
        | EQ => h
        | GT => qs (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qs (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qs (left k)}
          | EQ => hold {h}
          | GT => prev {qs (right, next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qs1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qs1 left, qs1 right)

fun qs2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qs2 (p1,k)
      | EQ => h
      | GT => qs2 (p2,k-n-1)
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\end{minipage}
\caption{Caption place holder}
\end{figure*}

@@

Importantly, one property of quickselect is that the partitioning of the list, even across recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  This independence allows the quickselect algorithm to be split into two functions:
one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qs1} in Figure~2c),
and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
and then used to accelerate many future lookups for any value of $k$.

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  After splitting, those parts of the \lang\ term in stage \bbone\ will end up in the precomputation, 
and those parts in stage \bbtwo\ will end up in the residual.
Typing rules for \lang ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

@@

\subsubsection{Dynamics}
\label{ssec:dynamics}

@@

@@

% would like some kind of intro here that anticipates all the complexity.
% The dynamics of \lang\ consists of three types of judgements: $\redonesym$ (stage-\bbone reduction), $\redtwosym$ (speculation), and $\reifysym$ (reification).

@@

\subsubsection {A Partial Evaluation System}
\label{sec:partialeval}

>>>>

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qsStaged (left k)}
          | EQ => hold {h}
          | GT => prev {qsStaged (right, 
                             next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qSelect1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qSelect1 left, qSelect1 right)

fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\end{minipage}
\caption{Caption place holder}
\end{figure*}

@@

Importantly, one property of quickselect is that the partitioning of the list, even across recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  This independence allows the quickselect algorithm to be split into two functions:
one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qSelect1} in Figure~2c),
and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
and then used to accelerate many future lookups for any value of $k$.

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  
After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
will end up in the stage \bbone\ function ({\tt qSelect1}), 
and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
Typing rules for \lang ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

@@

\subsubsection{Evaluation and Speculation}
\label{ssec:dynamics}

@@

The goal of the dynamics is to perform the above style of first-stage evaluation automatically.
It comprises judgments: $\redonesym$ (evaluation),
$\redtwosym$ (speculation), and $\reifysym$ (reification of substitutions into let bindings).

@@

@@

\subsubsection {Top-Level Evaluation}
\label{sec:partialeval}
